Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
done with VIII

Now we are going to show how from a given fonet
$\tuple{\gC,\linkingof\gC}$ we can construct a sequent proof $\Pi$ in
$\FOMLL$ such that $\fographof\Pi=\tuple{\gC,\linkingof\gC}$. In the
proof net literature, this operation is also called
\emph{sequentialization}. The basic idea behind our sequentialization
is to construct a propositional linked cograph, called the \bfit{frame}~\cite{hughes:unifn}
of $\gC$, in which the dependencies between the binders in $\gC$ are
encoded as links. Then we can apply the \emph{splitting tensor
theorem} to the frame, and then reconstruct the sequent proof $\Pi$.
\lutz{if the proof of thm \ref{thm:MLS1->fonet} is verified, we can delete the frame-def here}

More formally, let $\Gamma$ be a sequent with $\graphof\Gamma=\gC$, to which we
exhaustively apply the following
subformula rewriting steps, to obtain a sequent $\frameof\Gamma$:
\begin{enumerate}
\item {\bf Encode dependencies as fresh links.} For each dependency
  $\pair{\single x,\single y}$ in $\gC$, with corresponding
  subformulas $\exists x A$ and $\forall y B$ in $\Gamma$, we pick a
  fresh (nullary) predicate symbol $q$, and then replace $\exists
  x. A$ by $q \cand \exists x. A$, and replace $\forall y. B$ by
  $\dual q \cor \forall y. B$.
\item {\bf Erase quantifiers.} After step 1, remove all the
  quantifiers, i.e., replace $\exists x.A$ by $A$ and replace $\forall
  y.B$ by $B$ everywhere.
\item {\bf Simplify atoms.} After step 2, replace every predicate
  $p(t_1 \cdots t_n)$ (resp. $\dual p(t_1 \cdots t_n)$) with a nullary
  predicate symbol $p$ (resp.~$\dual p$)
\end{enumerate}
The $\linkingof{\frameof\Gamma}$ consists of the pairs induced by
$\linkingof\gC$ and the new pairs $\set{q,\dual q}$ introduced in
step~1 above. We call $\frameof\Gamma$ the \bfit{frame} of $\Gamma$ and we define the \bfit{frame} of $\gC$, denoted
$\frameof\gC$, as
$\tuple{\graphof{\frameof\Gamma},\linkingof{\frameof\Gamma}}$, and we immediately have the following:

\begin{lemma}\label{lem:frame}
  A linked fograph $\gC$ induces a bimatching iff its frame $\frameof\gC$ has an induced bimatching.
\end{lemma}

@@

The purpose of introducing the frame is the following theorem.

\begin{thm}
  \label{thm:splitting}
  Let $\Gamma$ be a propositional sequent containing only atoms
  and $\vlan$-formulas, and $\linkingof\Gamma$ be a
  linking for $\graphof\Gamma$. If $\tuple{
    \graphof\Gamma,\linkingof\Gamma}$ does not induce a bimatching then it is splittable.
\end{thm}

This is the well-know splitting-tensor-theorem
\cite{girard:87,danos:regnier:89}, adapted for the presence of
$\mix$. In the setting of linked cographs, it has first been proved by
Retor\'e~\cite{retore:03,retore:99}. We use it now for our sequentialization:

@@

\begin{proof}
  Let $\dsubstof\gC$ be the dualizer of $\tuple{\gC,\linkingof\gC}$.  We
  proceed by induction on the size of $\Gamma$ (i.e., the number of
  symbols in it, without counting the commas). If $\Gamma$ contains a
  formula with $\vlor$-root, or a formula $\forall x.A$, we can
  immediately apply the $\vlor$-rule or the $\forall$-rule of $\FOMLL$
  and proceed by induction hypothesis. If $\Gamma$ contains a formula
  $\exists x.A$ such that the corresponding binder $\single x$ in $\gC$
  has no dependency, then we can apply the $\exists$-rule, choosing the
  term $t$ as determined by $\dsubstof\gC$, and proceed by induction
  hypothesis.  Hence, we can now assume that $\Gamma$ contains only
  atoms, $\vlan$-formulas, or formulas of shape $\exists x.A$, where
  the vertex $\single x$ has dependencies. Then the frame
  $\tuple{\graphof{\frameof\Gamma},\linkingof{\frameof\Gamma}}$ does
  not induce a bimatching and contains only atoms and
  $\vlan$-formulas, and is therefore splittable. If it is mixed, then
  we can apply the $\mix$-rule to $\Gamma$ and apply the induction
  hypothesis to the two components. If it is not mixed then there must
  be a splitting tensor. If the splitting $\vlan$ is already in
  $\Gamma$, then we can apply the $\vlan$-rule and proceed by
  induction hypothesis on the two branches. However, if
  $\frameof\Gamma$ is not mixed and all splitting tensors are
  $\vlan$-formulas introduced in step~1 of the frame construction,
  then we get a contradiction as in that case there must be a $\vlor$- or $\forall$-formula in $\Gamma$.
  \lutz{can anyone give a good argument here?}
\end{proof}

@@

\begin{proof}
  We apply Theorem~\ref{thm:fonet->MLL1} to obtain a sequent
  proof~$\Pi$ of $\sqn C$ with
  $\fographof\Pi=\tuple{\gC,\linkingof\gC}$. Then we apply
  Lemma~\ref{lem:MLL1->MLS1}, observing that the translation from
  $\FOMLL$ to $\FOMLS$ preserves linking and dualizer.
\end{proof}

>>>>

Now we are going to show how from a given fonet
$\tuple{\gC,\linkingof\gC}$ we can construct a sequent proof $\Pi$ in
$\FOMLL$ such that $\fographof\Pi=\tuple{\gC,\linkingof\gC}$. In the
proof net literature, this operation is also called
\emph{sequentialization}. The basic idea behind our sequentialization
is to use the frame of $\gC$, to which  we can apply the \emph{splitting tensor
theorem}, and then reconstruct the sequent proof $\Pi$.

@@

\begin{thm}
  \label{thm:splitting}
  Let $\Gamma$ be a $\fff$-free propositional sequent containing only atoms
  and $\vlan$-formulas, and $\linkingof\Gamma$ be a
  linking for $\graphof\Gamma$. If $\tuple{
    \graphof\Gamma,\linkingof\Gamma}$ does not induce a bimatching then it is splittable.
\end{thm}

This is the well-known splitting-tensor-theorem
\cite{girard:87,danos:regnier:89}, adapted for the presence of
$\mix$. In the setting of linked cographs, it has first been proved by
Retor\'e~\cite{retore:03,retore:99} and then rediscovered by Hughes~\cite{hughes:pws}.
We use it now for our sequentialization:

@@

\begin{proof}
  Let $\dsubstof\gC$ be the dualizer of $\tuple{\gC,\linkingof\gC}$.  We
  proceed by induction on the size of $\Gamma$ (i.e., the number of
  symbols in it, without counting the commas). If $\Gamma$ contains a
  formula with $\vlor$-root, or a formula $\forall x.A$, we can
  immediately apply the $\vlor$-rule or the $\forall$-rule of $\FOMLL$
  and proceed by induction hypothesis. If $\Gamma$ contains a formula
  $\exists x.A$ such that the corresponding binder $\single x$ in $\gC$
  has no dependency, then we can apply the $\exists$-rule, choosing the
  term $t$ as determined by $\dsubstof\gC$, and proceed by induction
  hypothesis.  Hence, we can now assume that $\Gamma$ contains only
  atoms, $\vlan$-formulas, or formulas of shape $\exists x.A$, where
  the vertex $\single x$ has dependencies. Then the frame
  $\tuple{\graphof{\frameof\Gamma},\linkingof{\frameof\Gamma}}$ does
  not induce a bimatching and contains only atoms and
  $\vlan$-formulas, and is therefore splittable. If it is mixed, then
  we can apply the $\mix$-rule to $\Gamma$ and apply the induction
  hypothesis to the two components. If it is not mixed then there must
  be a splitting tensor. If the splitting $\vlan$ is already in
  $\Gamma$, then we can apply the $\vlan$-rule and proceed by
  induction hypothesis on the two branches. However, if
  $\frameof\Gamma$ is not mixed and all splitting tensors are
  $\vlan$-formulas introduced in step~1 of the frame construction,
  then we get a contradiction as in that case there must be a $\vlor$- or $\forall$-formula in $\Gamma$.
%  \lutz{can anyone give a good argument here?}
\end{proof}

@@

\begin{proof}
  We apply Theorem~\ref{thm:fonet->MLL1} to obtain a sequent
  proof~$\Pi$ of \hbox{$\sqn C$} with
  $\fographof\Pi=\tuple{\gC,\linkingof\gC}$. Then we apply
  Lemma~\ref{lem:MLL1->MLS1}, observing that the translation from
  $\FOMLL$ to $\FOMLS$ preserves linking and dualizer.
\end{proof}
